---
title: "Assessment"
author: "Paddy Kafeero"
date: "2026-01-30"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
#Load packages
source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")
```

Time of Origin: TIME OF DIAGNOSIS -- Same date as start of treatment
date. Treatment: Radiotherapy alone (0), Radiotherapy plus chemotherapy
(1)

```{r}
##Import the data
assignment <- rio::import("./assignment_data_2026.csv")
View(assignment)

##Labeling caterical values
assignment <- assignment %>% 
  mutate(
    status = factor(status, levels=c(0,1,2), labels = c("Censored","Disease relapse","Death without relapse")),
    trtgiven = factor(trtgiven, levels=c(0,1),labels=c("Radiotherapy only","Radiotherapy plus chemotherapy")),
    sex = factor(sex, levels = c(0,1), labels = c("Male","Female")),
    clinstg =factor(clinstg, levels = c(1,2), labels = c("Stage I","Stage II")),
    medwidsi = factor(medwidsi, levels = c(0,1,2), labels = c("None","Small","Large")),
    extranod = factor(extranod, levels = c(0,1), labels = c("No","Yes"))
  )

#View(assignment)
names(assignment)
```

#### Part 1: Descriptive analysis

Methods

a

Descriptive statistics for the the treatment strategies were described
as counts and percentages. Mean, standard deviation and range were used
for age and time by the treatment groups, whereas counts and percentages
were used for sex (Male/Female), status (Censored/Disease relapse/Death
without relapse). clinical stage at diagnosis (Stage I/Stage II), extent
of mediastinum involvement at diagnosis (Non/Small/Large) and
extra-nodal disease presence at diagnosis(Yes/No). Density plots were
used to investigate the distribution of age within the two treatment
groups, and ttests were used to assess presence of difference in time
and age by the two treatment groups.

b

A variable was created to contain the composite outcome of disease
relapse and death without relapse to be used for the analysis.

Kaplan-Meier estimates and plots were used to compare the survival
probabilities in the two treatment groups.

A log rank test for non-parametric methods was used to assess difference
in survival times between the two treatment strategies.

c

Aelan-Johansen estimates were used to ascertain the association between
treatment status ad the competing outcomes. These were also used to plot
cumulative incidence of the events by treatment group. A grays test was
used to check whether there was truly a difference in the incidence of
the events by the treatment strategies.

```{r}
##Summarising characteristics
summarytools::dfSummary(assignment) ##No missing data for any of the variables.
#assignment %>% group_by(trtgiven) %>% dfSummary()

tab1<-CreateTableOne(
  vars=c("age","sex","trtgiven","medwidsi","extranod","clinstg", "time","status"),
  strata="trtgiven",
  data=assignment,
  factorVars=c("sex","trtgiven","status","medwidsi","extranod","clinstg"),
  includeNA = FALSE,
  test = F,
  addOverall = T
)
print(tab1,showAllLevels = T)

##Age
assignment %>% group_by(trtgiven) %>% summarise(minage=min(age), medage=median(age), p75age=p75(age) ,maxage=max(age))


###Distribution of time and age: Do we assume normality for them?
ggplot(assignment, aes(x=age)) + geom_density()
ggplot(assignment, aes(x=time)) + geom_density()
##Ttest for time
t.test(assignment$time~assignment$trtgiven)
t.test(assignment$time~assignment$event)


###BY outcome
tab2<-CreateTableOne(
  vars=c("age","sex","trtgiven","medwidsi","extranod","clinstg", "time","status"),
  strata="status",
  data=assignment,
  factorVars=c("sex","trtgiven","medwidsi","extranod","clinstg"),
  includeNA = FALSE,
  test = F,
  addOverall = T
)
print(tab2,showAllLevels = T)



##Plots to see differences in treatment groups and outcome

```

Consider the composite outcome of disease relapse or death without
relapse (i.e. combining these two outcomes).

```{r}
##Summarising time to composite outcome: generate a binary variable for  an event
summary(assignment$time)
ggplot(assignment, aes(x=time)) + geom_density()

##Creating a composite outcome
assignment$event <- ifelse(assignment$status==0,0,1)

##Using KM
km_assign <- survfit(Surv(time, event)~trtgiven, data=assignment)
summary(km_assign)
summary(km_assign, censored = T)

plot(km_assign, lty = c(1,2), col = c("navyblue","maroon"),
     xlab="Time since diagnosis (months)",
     ylab="Survival probability"     )
title(main="Estimated survival probability plot",font=2)
legend("topright",c("Radiotherapy alone","Radio + Chemo"),
col=c("navyblue","maroon"), lwd = 2,lty = c(1,2))

##Fitting an weibul model
# weib_assign <-  mexhaz(Surv(time, event)~as.factor(trtgiven),base="weibull",data=assignment)
# summary(weib_assign) ##from this data we see radio+chemo people have HR 28% lower risk of the outcome. We are not assuming a constant rate here. This model can fit the data well based on the p-value for logRHO.cat("From the logRho value, we see that there is sufficient evidence to indicate that the survival between the two groups is different; with logRho =", 0.305864, "and the p-value < 0.0001 ", )

##Log rank test to check difference in survival within groups
survdiff(Surv(time, event)~trtgiven, data=assignment) ##
library(survminer)
ggsurvplot(km_assign, conf.int = T, pval = T, main="KM plots for treatment strategies", risk.table = TRUE, risk.table.clean=TRUE, risk.table.height=0.25)

```

(c) Consider disease relapse and death without relapse as separate
    events.

```{r}
#Summarising the time to disease relapse
summary(assignment$time[assignment$status==1]) ##minimum follow up is 0.003 months and max is 19.8 months
summary(assignment$time[assignment$status==2])

#For relapse
aataylor_relapse <- survfit(Surv(time, status==1) ~ trtgiven, data=assignment) ##Disease relapse
summary(aataylor_relapse)
survdiff(Surv(time, status==1) ~ trtgiven, data=assignment)
# For death without relapse
survfit(Surv(time, status == 2) ~ trtgiven, data = df)
survdiff(Surv(time, status==2) ~ trtgiven, data=assignment)


#Association between treatment status and the competing outcomes using non-parametric
Surv(time=assignment$time,event=assignment$status)
nonp_compet <- survfit(Surv(time, as.factor(status)) ~trtgiven, data=assignment) ##fitting the Aaelen Johnson estimator for cumulative incidence
summary(nonp_compet)

#plot(nonp_compet, lwd=2,col=c("lightblue","maroon"))
#legend("topleft",c("Disease relapse","Death without relapse"),
#lwd=2,col=c("lightblue","maroon"))

plot(nonp_compet,lwd=2,lty=c(1,2,1,2),col=c("navyblue","navyblue","maroon","maroon"),
     xlab = "Time from diagnosi",
     ylab = "Cumulative Incidence")
legend("bottomright",c("Radio, Disease relapse","Radio+Chemo, Disease relapse", "Radio, Death without relapse","Radio+Chemo, Death without relapse"), 
       lwd=2,lty=c(1,2,1,2),col=c("navyblue","navyblue","maroon","maroon"))

##using Grays test to test the NULL that there is no difference in the cumulative incidences by treatment given. The status variable should be left as a numeric variable and not changed to a factor
cuminc(ftime=assignment$time, fstatus=assignment$status, group=assignment$trtgiven)

```

Results

a

Out of the 845 participants, 71.1%(n=601) received radiotherapy alone.
Participants receiving radiotherapy alone were on average older than
those receiving radiotherapy plus chemotherapy(mean=36.02 vs 33.81
years). Participants

The average time to censoring, disease relapse or death without relapse
was longer in the intervention group (14.08 months) compared to the
radio alone group (12.33 months).

For both treatment groups, most patients were diagnosed with stage 2
clinical diagnosis (339[56.4%] for radio and 215[88.1] for radio plus
chemo), while most patients include in the study were diagnosed with
stage 2 diagnosis (65.6%). More participants receiving intervention had
extranodal disease at diagnosis (19.7) compared to those receiving radio
alone (4.8%). Most of the patients included in the study did not have
mediastnum involvement at diagnosis (53.7%, n=454), but a higher potion
of participants experiencing large mediastnum involvement were in the
intervention group (36.5% vs 3.8%) while the control group had more with
small mediastnum involvement (34.1% vs 30.4%).

On average, participants who died without relapse were older (mean
=47.85 years) than those who experienced the disease relapse (mean =
37.66 years), but this could be because older participants were more
fatally affected by the intervention procedures than the younger ones.
Additionally, participants who experienced disease relapse received it a
much earlier time (mean=2.65) than those who experienced death without
relapse (mean=12.37 months) indicating earlier experience of the
condition regardless of the treatment strategy used. Gender was balanced
within the three groups of status with slightly more men experiencing
death without relapse than women (n=84 vs n=46).

At ttest to investigate whether the difference is time between the two
treatment groups indicates evidence (p-value=0.009) of a difference
between the two groups.

Age was positively skewed, indicating that most participants in the
study were younger

b

Overall, the minimum time to event is 0.003 months and the maximum time
to event is 35.9 months. On average participants had experieced the
composite outcome by 12.8 months. From the survival plots, we notice
that the survival plots are crossing within the first 1 month with the
intervention group having steep fall in survival. however after 1 month,
the intervention group has higher suurvival and stabilises at around 25
months indicating lower risk of experiencing the event in the
intervention group. The control group survival plateus around 3 months
later. In general, the curves are very similar. The CI are wider at the
end of the cures becausee we have few events. 50% of the particpants had
experiencd the eent by 20 months, wheres less than 50% of the ppts In
the intervention group experienced the event.

For the two treatments, we have strong evidence for difference in time
to the event. p-value=0.005

c

For disease relapse, the minimum time to the event is 0.003 months
whereas the maximum time to the event is 19.8 months.

For death without relapse, minimum time to event is 0.6 months whereas
the maximum time to event is 29 months.

From the summary of the time we see that participants are experiencing
disease relapse way earlier than death without relapse, but also stop
getting event earlier than the latter.

Aelan-Johansen plot shows that a high incidence of disease relapse
within the first 5 months since diagnosis for both treatment strategies
and these stabilize at around 20 months. No more events are experienced
after 20 months. For death without relapse however, the plot shows a
steady increase in the cases for both groups, with the intervention
group plateauing earlier than the radio alone group.

From the Grays test: The null is that there is no difference in
cumulative incidence of the two treatment groups!

Disease relapse: Our p-value (0.0008) shows strong evidence of
differences in cumulative difference between the two groups.

Death without relapse: No evidence of a difference between the two
groups (p-value = 0.3740).

#### Part 2: Investigating conditional associations: composite outcome a) Cox regression analyses

Methods

A variable was created to contain the composite outcome of disease
relapse and death without relapse to beused for the analysis. A cox
proportional hazard model was created to check for association of
treatment strategy with the composite outcome. An adjusted cox model was
run to check for the association of the treatment with the outcome
conditional on other covariates in the dataset. Survival curves were
plotted for participants taking either of the treatment strategies and
conditional survival curves up to 12 month obtained. Estimated risks of
hazard up to 12 months were also obtained by subtracting the conditional
survival probability from 1.

```{r}
##Without adjustment
assignment$event <- ifelse(assignment$status==0,0,1)
table(assignment$event, assignment$status)
cox_p2 <- coxph(Surv(time, event)~trtgiven, data = assignment)
summary(cox_p2) ##The HR and the p-values for this model are similar to the Weibull distribution model run on the same data. p-value = 0.005 & HR=0.7254
  #plot values using the cox model and compare to the KM plot
  assign_p2_survfit=survfit(cox_p2,newdata=data.frame(trtgiven=c(0,1)))
plot(assign_p2_survfit,conf.int = T,
     col=c("black","grey"),
     xlab="Time since diagnosis", ylab="Survival probability")
legend(x="bottomleft",c("Placebo","Active"),
       col=c("black","grey"),lty=1)

summary(pbc.survfit,censored = T,times = 5)

##The plot of the cox model and the KM plot are different. In the KM, we assume the proportional hazard is constant between the two groups over the period of study. The KM is non-parametric though.
plot(assign_p2_survfit,fun="cloglog",conf.int = T,
xlab="time (log scale)",ylab="log(-log S(t))",
col=c("black","grey")) ##checking the proportional hazards assumption, we confirm that the assumption is true given the parallel lines with no overlapping.

##With adjustment: Assuming NO ASSUMPTION OF ASSOCIATION
cox_p2_adj <- coxph(Surv(time, event)~factor(trtgiven)+age+sex+clinstg+factor(medwidsi)+factor(extranod), data = assignment, x=T)
summary(cox_p2_adj) ##how do we check the assumption that age is 


#Assumptions of the multivariable adjusted COx model.
#PH assumption: For the cox model, we use the schoenfield residuals method
sch.resid_p2a<-cox.zph(cox_p2_adj, transform = 'identity')
sch.resid_p2a
par(mfrow=c(2,2))
plot(sch.resid_p2a,col="red",lwd=2)

##We can assess whether the continuous variables should be transformed using Martingale residuals. Check whether functional form of age is appropriate for our model
#Martingale plot foor age
p2.cox.noage<-coxph(Surv(time=time,event=event)~trtgiven+sex+clinstg+factor(medwidsi)+factor(extranod),data=assignment)
mgale.res.noage<-resid(p2.cox.noage,type="martingale")
plot(assignment$age,mgale.res.noage,
     xlab="Age",ylab="Martingale residuals")
lines(lowess(assignment$age,mgale.res.noage),col="red",lwd=2)
abline(h=0,lwd=2,col="grey")
#Martingale residual plot for age:including age in Coxmodel
p2.cox.age<-coxph(Surv(time=time,event=event)~trtgiven+age+sex+clinstg+factor(medwidsi)+factor(extranod), data=assignment)
mgale.res.age<-resid(p2.cox.age,type="martingale")

plot(assignment$age, mgale.res.age, xlab="Age", ylab="Martingale residuals")
lines(lowess(assignment$age, mgale.res.age),col="red",lwd=2)
abline(h=0,lwd=2,col="grey")


##The martingale including age is fine, we notice that age is in its correct functional format to be used in the model
```

b)  Conditional survival probabilities

```{r}
##Create a dataset with radio therapy only and then compare a new dataset
p2b_data_radio <- expand.grid(
  trtgiven = 0, ##Radiotherapy only
  clinstg=1,
  age=50,
  sex=1, ##1 for Female
  medwidsi=1,
  extranod=1
)



#Getting estimated survival curves
p2b_radio <- survfit(cox_p2_adj, newdata = p2b_data_radio)
summary(p2b_radio)

plot(p2b_radio)

##Radiotherapy+Chemo
p2b_data_radiochem <- p2b_data_radio
p2b_data_radiochem$trtgiven <- 1 ##Radiotherapy + Chemotherapy
p2b_radiochemo <- survfit(cox_p2_adj, newdata = p2b_data_radiochem)
plot(p2b_radiochemo)

names(p2b_radiochemo)

##Plots comparing the probabilities
plot(p2b_radio, conf.int = F, col = "black",
     xlab = "Time since diagnosis (months)",
     ylab = "Survival probability",
     main = "Estimated Survival curves plots under two the two treatment strategies ")
lines(p2b_radiochemo, conf.int = F, col = "grey")
legend("topright", c("Radio", "Radio+Chemo"), col=c("black","grey"), lwd=2)


##############################################################
#p2b_2 <- survfit(Surv(time, event)~trtgiven, data = assignment)
#plot(p2b_2)

##At 12 months
#summary(p2b_2, times=12)
p2b_riskradio <- 1-summary(p2b_radio, times=12)$surv
p2b_riskradiochemo <- 1-summary(p2b_radiochemo, times=12)$surv


```

Results

a

Without adjustment

The HR for radio plus chemo is 0.7254 (95% CI 0.58 to 0.907). This
indicates that the radio plus chemo group is associated with a 27% lower
risk of the composite outcome compared to the radiotherapy alone group.
There is also some strong evidence of this association (p-value=0.005).

With adjustment

The HR for radio plus chemo is 0.6525 (95% CI 0.5043 to 0.8442), for age
1.033(95% CI 1.027 to 1.04), sex 0.8546 (95% CI 0.7003 to 1.0429), for
clinical stage diagnosis 1.4481 (95% CI 1.1491 to 1.8249), for Small
mediastinum involvemnt at diagnosis 0.82(95% CI 0.6365 to 1.0564), for
large mediastinum involvement 1.0184 (95% CI 0.7015 to 1.4783) and for
extranodal disease 1.1829 (95% CI 0.8324 to 1.6809).

From this, we can say that theintervention group sis estimated to have a
hazard for the composite ooutcome that is 35% lower than the
radiotherapy alone group, conditional on age, sex, clinical stage,
mediastnum involvement and extranodal disease. For age, we can say that
a unit increase in age is associated with a 3% increase in hazard
conditional on other covariates(we have very strong evidence). Females
have a 15% lower hazard compared to the males (no evidence). Patients
with stage 2 clinical stage at diagnosis have a 45% higher risk than
stage 1, conditional on other covariates (strong evidence). Patients
with small extent of mediastinum involvement at diagnosis are estimated
to have a hazard that is 18% lower than those without mediastinum
involvement, conditional on other covariates (no evidence). On the
contrary patients with large extent of mediastinum involvement at
diagnosis are estimated to have a hazard that is 2% higher than those
without mediastinum involvement, conditional on other covariates (no
evidence). Participants will extranodal disease have an estimated hazard
of of the composite outcome that is 18% higher that those without,
conditional on the other covariates, and there is no evidence for this
association though.

b

Figrure: Estimated conditional survival curves for the two treatment
strategies

Estimated risk of disease relapse/death without relapse for radiotherapy
alone is 0.5276779 compared to 0.3870146 of the radiotherapy plus
chemotherapy group. We can see that for the treatment groups, based on
the composite outcome, intervention group shows that it is more
beneficial than the radio alone group.

#### Part 3: Investigating conditional associations: competing events a) Cause-specific cox

Methods

Cause-specific proportional hazards models were used to compare the
hazard of disease relapse and death without relapse between the
radiotherapy alone and the radiotherapy plus chemotherapy groups.
Additionally, they were also used to generate cumulative incidence
curves and the cumulative incidence at 12 months for a patient that ...
An Aalen-Johnson cumulative incidence curve was generated to compare the
cumulative incidence for the two events when considering other
covariates.

The grays test was used to test the null hypothesis of whether there was
a difference in the cumulative incidences for disease relapse and death
without relapse between the two groups.

```{r}
##Looking at the survival times
Surv(time=assignment$time,event=as.factor(assignment$status))[1:10]

##Lets try too compare the cumulative incidences and their plots for the competing groups
#Estimate the cumulative incidences using the Aalen-Johansen estimator
cum_p3 <- survfit(Surv(time, as.factor(status)) ~trtgiven, data=assignment)
#plot the cumulative incidences
# plot(cum_p3, lwd=2,col=c("black","grey"),
#      xlab="Time since randomisation (years)",
#      ylab="Cumulative incidence")
# legend("topleft",c("Disease relape","Death without relapse"),
#        lwd=2,col=c("black","grey"))

plot(cum_p3,lwd=2,lty=c(1,2,1,2),col=c("navyblue","navyblue","maroon","maroon"),
     xlab = "Time from diagnosi",
     ylab = "Cumulative Incidence")
legend("topleft",c("Radio, Disease relapse","Radio+Chemo, Disease relapse", "Radio, Death without relapse","Radio+Chemo, Death without relapse"),
       lwd=2,lty=c(1,2,1,2),col=c("navyblue","navyblue","maroon","maroon"))

#Without adjustment
cox_p3 <- coxph(Surv(time, as.factor(status))~trtgiven, data = assignment, id=id)
summary(cox_p3)
  
#With adjustment
cox_p3_adj <- coxph(Surv(time, as.factor(status))~trtgiven+age+sex+clinstg+factor(medwidsi)+extranod, data = assignment, id=id)
summary(cox_p3_adj)



```

b)  Cumulative incidences

```{r}
#Getting estimated survival
#Radiotherapy alone
p3b_radio <- survfit(cox_p3_adj, newdata = p2b_data_radio)
summary(p3b_radio)

plot(p3b_radio)

plot(p3b_radio, ylab="Cumulative incidence",
     xlab = "Times from diagnosis (months)",
     main="Cumulative incidence curves for the specific events with radiotherapy alone",
     lwd = 2, lty = c(1,2))
legend("bottomright", c("Disease Relapse", "DeathNoRelapse"),
       lwd=2,
       lty=c(1,2)
       )


##Radiotherapy+Chemo
p3b_data_radiochem <- p3b_radio
p3b_data_radiochem$trtgiven <- 1 ##Radiotherapy + Chemotherapy
p3b_radiochemo <- survfit(cox_p3_adj, newdata = p2b_data_radiochem)
plot(p3b_radiochemo)

plot(p3b_radiochemo, ylab="Cumulative incidence",
     xlab = "Times from diagnosis (months)",
     main="Cumulative incidence curves for the specific events with radiotherapy plus chemotherapy",
     lwd = 2, lty = c(1,2))
legend("bottomright", c("Disease Relapse", "DeathNoRelapse"),
       lwd=2,
       lty=c(1,2)
       )

###COMBINED PLOT FOR CUMULATIVE INCIDENCES
##Plots comparing the probabilities
plot(p3b_radio, conf.int = F, col = c("black","grey"), lwd=2, lty = 1,
     xlab = "Timme in months",
     ylab = "Cumulative Incidence",
     main = "Estimated Survival curves plots under two the two treatment strategies ", ylim = c(0,0.6))
lines(p3b_radiochemo, conf.int = F, col = c("black","grey"), lwd=2, lty = 2)
legend("bottomright",c("Radio, Disease relapse","Radio+Chemo, Disease relapse", "Radio, Death without relapse","Radio+Chemo, Death without relapse"), lwd=2,lty=c(1,2,1,2),col=c("black","black","grey","grey"))

plot(p3b_radio, conf.int = F, col = c("black","grey"), lwd=2, lty = 1,
     xlab = "Timme in months",
     ylab = "Cumulative Incidence",
     main = "Estimated Survival curves plots under two the two treatment strategies ", ylim = c(0,0.6))
lines(p3b_radiochemo, conf.int = F, col = c("black","grey"), lwd=2, lty = 2)
legend("topleft",c("RO: DR","RC: DR", "RO: DWR","RC: DWR"), lwd=2,lty=c(1,2,1,2),col=c("black","black","grey","grey"), text.width = 6)



#legend("topright", c("DiseaseRelapse", "DeathNoRelapse"), col=c("black","grey"), lwd=2)


##CUMULATIVE INCIDENCES At 12 months
summary(p3b_radio, times=c(1,12,24))
summary(p3b_radiochemo, times=c(1,12,24))
##using Grays test to test the NULL that there is no difference in the cumulative incidences by treatment given. The status variable should be left as a numeric variable and not changed to a factor
cuminc(ftime=assignment$time, fstatus=assignment$status, group=assignment$trtgiven)
##from the test, we se ethat we have evidence against the null. For Disease Relapse, cumulative incidence between the treatment groups is different
#For Death without relapse: p=0.37 indicating that we have no evidence of diffences in cumulative  incidence for the treament groups. We do not reject the null.
```

Results

a)  

Unadjusted

For the disease relapse: the HR is 0.620 (95% CI 0.467-0.822). This
means that at a given time since diagnosis, the hazard of disease
relapse among people receiving radiotherapy+chemo is 0.62 times the
hazard of those receiving radiotherapy alone. The 95% CI does not
contain the null value, indicating that there is evidence that this
cause-specific hazard ratio is not 1. For death without relapse: the
cause-specific HR is 0.98 (95% CI 0.679 - 1.416). This means that at a
given time since diagnosis, the hazard of death without relapse among
those receiving the intervention is 0.98 time that of those receiving
radiotherapy alone. However, since the 95% CI contains the null value,
we do not have enough evidence against the null that the cause specific
hazard ratio is 1.

Adjusted

For disease relapse: the estimated HR for treatment is 0.493(95% CI
0.3523 to 0.6909), for age 1.016(95% CI 1.008 to 1.024), sex 0.971 (95%
CI 0.764 to 1.233), for clinical stage diagnosis 1.570 (95% CI 1.186 to
2.077), for Small mediastinum involvemnt at diagnosis 0.749(0.557 to
1.007), for large mediastinum involvement 1.133 (95% CI 0.735 to 1.748)
and for extranodal disease 1.374 (95% CI 0.901 to 2.097). From this, we
can say that theintervention group sis estimated to have a hazard for
disease relapse that is 51% lower than the radiotherapy alone group,
conditional on age, sex, clinical stage, mediastnum involvement and
extranodal disease. For age, we can say that a unit increase in age is
associated with a 2% increase in hazard of disease relapse(we have some
evidence). Females have a 3% lower hazard of disease relapse compared to
the males (no evidence). Patients with stage 2 clinical stage at
diagnosis have a 57% higher risk of disease relapse that stage 1,
conditional on other covariates (string evidence). Patients with small
extent of mediastinum involvement at diagnosis are estimated to have a
hazard of disease relapse that is 26% lower than those without
mediastinum involvement, conditional on other covariates. nOn the
contrary patients with large extent of mediastinum involvement at
diagnosis are estimated to have a hazard of disease relapse that is 13%
higher than those without mediastinum involvement, conditional on other
covariates. Participants will extranodal disease have an estimated
hazard of disease relapse that iis 37 % higher that those without,
conditional on the other covariates, and there is no evidence for this
association though.

For death without relapse: the estimated HR for treatment is 1.082(95%
CI 0.728 to 1.608), for age 1.081(95% CI 1.068 to 1.094), sex 0.614 (95%
CI 0.414 to 0.910), for clinical stage diagnosis 1.381 (95% CI 0.898 to
2.124), for Small mediastinum involvemnt at diagnosis 0.987(0.597 to
1.635), for large mediastinum involvement 0.764 (95% CI 0.353 to 1.653)
and for extranodal disease 0.821 (95% CI 0.436 to 1.543).

From this, we can say that theintervention group is estimated to have a
hazard for death without relapse that is 8% higher than the radiotherapy
alone group, conditional on age, sex, clinical stage, mediastnum
involvement and extranodal disease (no evidence). For age, we can say
that a unit increase in age is associated with an 18% increase in hazard
of death without relapse(we have some evidence). Females have a 39%
lower hazard of death without relapse compared to the males (we have
evidence). Patients with stage 2 clinical stage at diagnosis have a 38%
higher risk of death without relapse that stage 1, conditional on other
covariates (no evidence). Patients with small extent of mediastinum
involvement at diagnosis are estimated to have a hazard of death without
relapse that is 2% lower than those without mediastinum involvement,
conditional on other covariates. On the contrary patients with large
extent of mediastinum involvement at diagnosis are estimated to have a
hazard of death without relapse that is 24% lower than those without
mediastinum involvement, conditional on other covariates (no evidence).
Participants will extranodal disease have an estimated hazard of death
without relapse that is 18% lower than those without, conditional on the
other covariates, and there is no evidence for this association though
(asically difference is due to random chance!).

b

Figure title: Cumulative incidences of Disease Relapse and Death without
relapse by treatment given using cause-specific proportional hazards
model

We see from the curves that for disease relapse has high incidence
within the first 5 months of diagnosis for both treatment strategies
indicating high risk of events within the first 5 months. However, this
rate slows down up to around 15 months and platues at 20 months where we
have no more events in both groups.

For death without relapse,

Additionally, we can see that the cause-specific proportional hazards
assumption is reasonable for both disease relapse and death without
relapse. However, when we compare this to the Aalen-Johnson estimates
plot for cumulative incidences, we see clear differences where the PH
assumption doesn't stand for the deaths without relapse event.

Cumulative incidence

Radiotherapy alone

CI for disease relapse is 37.5% (0.375)

CI for deaths without relapse is 6.6%(0.065798)

Radiotherapy plus chemotherapy

CI for disease relapse is 21.1%(0.2110)

CI for death without relapse is 13.7%(0..1370)

The grays test for the cumulative incidence of the events by treatment
group tell us:

For Disease relapse: We have p-value of 0.0007 indicating very strong
evidence against the null that the cumulative incidence between the two
treatment strategies is the same.

For Death with out relapse: We have a p-value of 0.3739 indicating very
weak evidence again the null. We can then conclude that there is no
evidence of a difference in cumulative incidence of death without
relapse between the two treatment groups.

#### Part 4: Estimating a marginal treatment effect

Methods To get the marginal risk for the two two treatment strategies,
two separate datasets based on the original data were creating with all
observations in each of the datasets having either the radiotherapy plus
chemotherapy or the radiotherapy alone. Using the cox adjusted model,
predicted survival probabilities at 12 months were obtained for each
observation and these were used to obtain the average risk at that
timepoint.

Marginal treatment risk were obtained by generating the Restricted Mean
Survival Time for the two treatment groups, and by generating the risk
ratios for the two groups. A milestone time of 12 months was chosen due
to the fact that chemotherapy in cancer treatment is considered to have
harmful effects within the short term of intervention and the effects
start to get realised after about a year on treatment. Comparing the
RMST and risk ratio values at 6, 12 and 24 months also identified the 12
months as an optimal milestone time point for assess and determine
treatment effects for the two groups. The individual group risks were
obtained from the Kaplan Meier estimates for event-free survival at the
specific time points.

Composite outcome a) Marginal risks

```{r}
#Based on the model in 2b: cox_p2_adj
assignment$trtgiven <- as.factor(assignment$trtgiven)
adjsurv<-adjustedsurv(data=assignment,
                      variable="trtgiven",
                      ev_time="time",
                      event="event",
                      method="direct",
                      outcome_model=cox_p2_adj,
                      conf_int=TRUE,
                      bootstrap=F,times=12)
# plot(adjsurv,conf_int=T,custom_colors=c("black","grey"),  ###this plot only works if we do not speccify a time, it is meant for the marginal risk all through the study period.
#      xlab="Time(months)",
#      legend.title="Treatment",
#      ylim=c(0,1))

#Risks at 12 months
adjsurv$ate_object$meanRisk

##risk diff at time 12
adjsurv$ate_object$diffRisk

#risk ratio at time 12
adjsurv$ate_object$ratioRisk




###MANUALLY
#'byhand' implementation
#--
#create dataset in which trtgiven is set to 1 for everyone
p3_data_radiochemo<-assignment
p3_data_radiochemo$trtgiven<-1
#create dataset in which trtgiven is set to 0 for everyone
p3_data_radio<-assignment
p3_data_radio$trtgiven<-0

#predicted risk up to time 12 for each individual using cox_p2_adj
#Using predictSurvProb from pec package- neater
risk.radiochemo<-1-predictSurvProb(cox_p2_adj,newdata=p3_data_radiochemo,times=12)
risk.radio<-1-predictSurvProb(cox_p2_adj,newdata=p3_data_radio,times=12)
cbind(risk.radiochemo, risk.radio)[1:5]
#meansurvivalprobabilityateachtime,averagingoverallindividuals
riskmean.radiochemo<-mean(risk.radiochemo)
riskmean.radio<-mean(risk.radio)


```

| Estimand | Estimate | 95% CI |
|----|----|----|
| Marginal risk up to 12 years: Radiotherapy+Chemotherapy | 0.317 | 0.266 to 0.368 |
| Marginal risk up to 12 years: Radiotherapy Only | 0.431 | 0.266 to 0.368 |
| Marginal risk difference | -0.114 | -0.178 to -0.051 |

b)  Marginal treatment risks

```{r}
##For this part, we need to choose a MILESTONE TIMEPOINT for which we can get a  risk ratio and a restriccted mean difference. Get the milestone time and give justification for it. We can look at the risk difference p-values and the restricted mean p-values and compare the to, to see if we have an appropriate value and to see if both measurements make sense.
#We shall have to first plot the KM graph to see
#Kaplan-Meier by treatment
km_p4<-survfit(Surv(time,event)~trtgiven,data=assignment)

plot(km_p4, ylab="Survival probability",
     xlab = "Times from diagnosis (months)",
     lwd = 2, lty = c(1,2))
legend("topright", c("Radio", "Radio+Chemo"),
       lwd=2,
       lty=c(1,2)
       )

#Radio plus Chem first has a rapid fall in the survival before they cross over at 

##Restricted mean diff: This method compares area under a curve at specific timeperiods and it is best used for cases like these where we observe a cross over.
print(km_p4,rmean=6)
rmst2(assignment$time, assignment$event, assignment$trtgiven, tau=6)

##Risk Ratio: We get the risks a 1 year and see. We get risks using the KM estimates (The survival probabilties)
#riskdiffsat8m
risk_radio<-1-summary(km_p4,times=c(6))$surv[1]
risk_radiochem<-1-summary(km_p4,times=c(6))$surv[2]

varrisk_radio<-summary(km_p4,times=c(6))$std.err[1]^2

varrisk_radiochem<-summary(km_p4,times=c(6))$std.err[2]^2
var.riskdiff<-varrisk_radiochem+varrisk_radio #Variance for the two independent times

lower.ci.riskdiff<-(risk_radiochem/risk_radio)-1.96*sqrt(var.riskdiff) ##Risk Ratio Lower CI
upper.ci.riskdiff<-(risk_radiochem/risk_radio)+1.96*sqrt(var.riskdiff) ##Risk Ratio Upper CI

risk_radiochem/risk_radio ##Risk Ratio

lower.ci.riskdiff
upper.ci.riskdiff

2*pnorm((risk_radiochem-risk_radio)/sqrt(var.riskdiff))#p-


##Working with a milestone time point of 1 year (12 months) allows for us to identify sufficient time for us to allow for accummulation of treatment benefit for the intervention.
#At 6 months, the risk ratio provides way more evidence of a treatment benefit than the RSMT and this is because the time-frame is inappropriately short and has not allowed given enough time for the treatment benefit to accumulate.
#At 24 months however, we see that the the risk ratio did not give any evidence of a treatment benefit (p-value=0.02), whereas the RSMT gave string evidence (p-value=0.001) of treatment evidence.



```

Results a) If everyone had been given radiotherapy alone, the expected
marginal probability of disease relapse/death without relapsse up to 12
months is 43.1%. On the otherhand, if everyone has been given
radiotherapy with chemotherapy the expected marginal probability up to
12 month sis 31.7%.

b)  

A Kaplan Meier plot shows a cross over between the treatment curves for
the Radiotherapy alone and the intervention group, showing initial harm
from the intervention for a period of about 1 month before this changes.

At 12 months, the RSMT difference is 1.211 months (95% CI; 0.538 -
1.884; p-value\<0.001) indicating that within 12 months of initiating
radio plus chemo, time to disease relapse/death without relapse is
delayed by about 1.2 months in the intervention group compared to the
radio alone group. The Risk Ratio: 0.699(95% CI; 0.629 - 0.770;
p-value=0.0002) indicates that within 12 months the risk of disease
relapse/death without relapse is lower by about 31% within the
intervention group compared to the radio alone group.

At this point, we check milestone times of 6, 12 and 24 months, and the
12 months time point seemed to be the most ideal point, because of the
fact that it produces relevant and significant results with both
measures used. At 6 months, the Risk ratio was working better because
the intervention group treatment effect was just being realised just
after a cross over at around 1 month. conversely, at the 24 months
period, only the RSMT gave evidence and this is because the treatment
effects by both groups were stabilising at this point, and few cases
were being obtained (Risk ratio and risk measurements are affected by
the number of cases later on in the study).

Summary

We see that with the cuase specific analysis, the intervention is good
for disease relapse but not for the death without relapse. Looking at
the cumulative incidence curves for the model with cause-specif shows
this.
