---
title: "Session7"
author: "Paddy Kafeero"
date: "2026-01-23"
output: html_document
---


```{r}
#Load packages
source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")
```


Question 1
```{r}
##Loading the data
load("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Survival_Analysis/survival_rfiles/aaatrial_2016.RData")
summary(aaatrial)
View(aaatrial)

aaatrial <- aaatrial %>%  select(id, group, dateran, aaadeath, alldeath, deathtype, timeout, dob)

aaatrial$group <-  factor(aaatrial$group, levels =c(0,1), labels = c("Control","Invited"))
aaatrial$aaadeath <-  factor(aaatrial$aaadeath, levels =c(0,1), labels = c("Censored/Non-AAA-death","AAA-death"))
aaatrial$alldeath <-  factor(aaatrial$alldeath, levels =c(0,1), labels = c("Censored","Death(any cause)"))
aaatrial$deathtype <-  factor(aaatrial$deathtype, levels =c(0,1,2), labels = c("Censored","non-AAA death","AAA-death"))

table(aaatrial$deathtype)
table(aaatrial$group)

##Followup in years using dateran and timeout
aaatrial$time <- as.numeric(difftime(aaatrial$timeout, aaatrial$dateran, units="days")) / 365.25
View(aaatrial)

##Cause specific cox model
cox.cs <- coxph(Surv(time, as.factor(deathtype)) ~ group, data=aaatrial, id=id)
summary(cox.cs)
  #form of the hazard models: h1(t|x) = ho1Exp(eB1x)

#equivalent way
cox.AAA<-coxph(Surv(time,deathtype=="AAA-death")~group,data=aaatrial)
summary(cox.AAA)
# cox.AAA<-coxph(Surv(time,aaadeath)~group,data=aaatrial)#same
cox.nonAAA<-coxph(Surv(time,deathtype=="non-AAA death")~group,data=aaatrial)
summary(cox.nonAAA)

#How the competing event is treated in fitting cause-specific Cox models
#People without the event of interested are not considered part of the risk set but rather as censored individuals

```
groupInvited_1:2     0.977: Hazard of non-AAA death is lowered by 2% if the person is invited vs when they are a control
groupInvited_1:3     0.669: Hazard of AAA death is lowered by 33% if the person is invited vs when they are a control

Question 2
```{r}
cinc.cs <- survfit(cox.cs,newdata = data.frame(group=c("Control","Invited")))
summary(cinc.cs, times=c(1,5,10))  ###surv fit will know that because death type is a factor, i am interested in cumulative incidence. group has two individuals


plot(cinc.cs,lwd=2,lty=c(1,2,1,2),col=c("black","black","grey","grey"))
legend("topleft",c("Control, Non-AAA-death","Invited, Non-AAA-death",
                   "Control, AAA-death","Invited, AAA-death"),
       lwd=2,lty=c(1,2,1,2),col=c("black","black","grey","grey"))
```
Cumulative incidence for AAA death at time 5: Control(0.1371); Invited(0.1343)
Cumulative incidence for non-AAA death at time 5: Control(0.00521); Invited(0.00349)


Question 3
```{r}
aaatrial$age_yrs <- as.numeric(difftime(aaatrial$dateran, aaatrial$dob, units="days")) / 365.25
View(aaatrial)

##Cause specific cox model
cox.cs_age <- coxph(Surv(time, as.factor(deathtype)) ~ group+age_yrs, data=aaatrial, id=id)
summary(cox.cs_age)


#Equivalent
cox.AAA_age<-coxph(Surv(time,deathtype=="AAA-death")~group+age_yrs,data=aaatrial)
summary(cox.AAA_age)
# cox.AAA<-coxph(Surv(time,aaadeath)~group,data=aaatrial)#same
cox.nonAAA_age<-coxph(Surv(time,deathtype=="non-AAA death")~group+age_yrs,data=aaatrial)
summary(cox.nonAAA_age)


```
AAA deaths: HRS
groupInvited    0.6703
age_yrs         1.1602: Given that an individual is alive at a specific time, an increase in age by 1 year increased the hazard of Non-aaa deaths by 16% keeping other factors constant.

Non-AAA deaths: HRs
groupInvited     0.978


Question 4
```{r}

cinc.cs_age <- survfit(cox.cs_age,newdata = data.frame(id=1:4, group=c("Control","Invited","Control","Invited"), age_yrs=c(65,65,70,70)))  ##creatng a dataframe for just 4 people

summary(cinc.cs_age, times=c(1,5,10))


```
Person 1: nonAAA (cum=0.0896); AAA(cum=0.002638)
Person 2: nonAAA (cum=0.0877); AAA(cum=0.001771)
Person 3: nonAAA (cum=0.143); AAA(cum=0.00538)
Person 4: nonAAA (cum=0.1401); AAA(cum=0.00362)



Question 5: Fine and Gray models (subdistribution hazard models)
```{r}
crr.AAA <- crr(aaatrial$time, aaatrial$deathtype, 
               cov1=aaatrial$group, failcode=2)  ###group and deathtype shouldnt be factors
summary(crr.AAA)


```
aaatrial$group1     0.671 This is a subdistribution hazard ratio.


Question 6
```{r}
cinc.fg.AAA <- predict.crr(crr.AAA, cov1 = c(0,1))
#Note:
#Column 1 is time
#Column 2 is cumulative incidence for AAA in group 0
#Column 2 is cumulative incidence for AAA in group 1
head(cinc.fg.AAA)

```



Question 7
```{r}
#i: Non-parametric method of getting cumulatice incidence. Not assuming any model assumptions

plot(cinc.cs)

#ii Cause specific: assuming proportional hazards
plot(cinc.cs)

#iii :Assuming proportionality of subdistribution hazards for AAA deaths
plot(x=cinc.fg.AAA[,1], y=cinc.fg.AAA[,2], lwd=2, type = "s")
plot(x=cinc.fg.AAA[,1], y=cinc.fg.AAA[,3], lwd=2, type = "s")





```



